Comparing Machine Learning algorithms for classifying cognitive states from fMRI data

### 2012-11-12 article/article.tex
review

%% The ``\maketitle'' command must be the first command after the
%% ``\begin{document}'' command. It prepares and prints the title block.
%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}
\label{sec:introduction}
\maketitle
%% Now the actual contents of first section
Human functional brain mapping is a field of study that has grown strongly in the past decades. This is not a surprise, because it is one of the few ways to look at brain activity of a living animal. There are several techniques available to scan for brain activity in a living human. Of these techniques, functional magnetic resonance imaging (fMRI) has become more and more popular because it has a very good spatial resolution (contrary to MEG/EEG), and it is non-invasive (contrary to PET). Brain activity is inferred from differences in the ratio between oxygenated and deoxygenated blood; when brain activity occurs, more oxygenated blood will flow to a brain region, and this change is detected by a machine. This is called the blood oxygen level dependent (BOLD) response \cite{sm:2012fMRI}.

@@

Traditionally, data from each voxel, or from a group of voxels, are analyzed with analysis of variance. Voxels are usually convoluted with a haemodynamic response function in an attempt to infer the spiking neural activity from the slowly building and dropping BOLD response \cite{he:2002na} (cf.~Figure~\ref{fig:bold_response}). In this setup, experimental conditions are the independent variables and a general linear model is fitted to each voxel, or group of voxels. By using this technique, voxels or brain areas can be identified that differ in their average BOLD response across conditions.

@@

\subsection{Dimensionality Reduction}
\label{sec:dimensionalityReduction}
%Useful background info: http://www.brainvoyager.com/bvqx/doc/UsersGuide/WebHelp/Content/MVPATools/MVPA_Basic_Concepts.htm
Typical input for the machine learning algorithms discussed in this paper consists of a number of feature vectors, one for each fMRI scan. Such a feature vector $\vec{x}$ contains measures for the $N$ features we use to represent that scan and is derived from our input data. In the most basic case, we could use the entire set of voxels belonging to a scan, resulting in a feature vector containing potentially (hundreds of) thousands of features.
In the case that we have a very large dimension of features combined with only a limited amount of scans, our classifier may have difficulty learning the right classification function. This problem is formally known as the \textit{Hughes effect}\cite{Hughes1054102} which states that the predictive power of a machine learning algorithm decreases as the dimensionality increases with a fixed number of training examples. To reduce the dimensionality of the input, several different methods were used.

@@

\textbf{\emph{ROI mapping.}} ROI mapping is a form of feature selection, wherein only a subset of the N-dimensional input space is used. As described in the introduction, ROIs are anatomically defined brain areas classified as relevant by an expert in the field. We used 7 of the 25 defined ROIs, labeled  'CALC' 'LIPL' 'LT' 'LTRIA' 'LOPER' 'LIPS' and  'LDLPFC' (see Figure~\ref{fig:fd}). These ROIs have previously been found to be most relevant in discriminating between the cognitive states of looking at sentences or pictures \cite{wa:2003betw}. To further reduce the number of features, average voxel values within these regions are taken as new input features. As with preprocessing, other higher-order features such as the standard deviation, kurtosis and skewness within ROIs did not increase predictability and were not included in the subsequent analysis. The disadvantage of using ROIs is that they require \textit{a priori} knowledge. When this knowledge is inaccurate, classification accuracy may not be optimal.

@@

\textbf{\emph{Temporal dimensionality reduction.}} The BOLD response peaks at approximately 5 seconds after a stimulus is presented (see Fig. 1). This means that the first scans after a stimulus may not represent activity related to this stimulus and should not be used for classification. To examine which scans provide maximal class information, we plotted the average BOLD activity per scan for all trials of all subjects, given a trial type and ROI. As can be seen in Figure~\ref{fig:boldroi}, scans 10 to 20 provide maximal information for the first stimulus, while scans 23 to 33 provide maximal information for the second stimulus. The other scans were excluded.      
\begin{figure}
	\centering
	\includegraphics[width=90mm]{figures/scan_hemodynamic_BOLD_roi}
  	\caption{Haemodynamic response for ROI "CALC". The red line depicts a sentence-picture trial (S-P) and the blue line depicts a picture-sentence trial (P-S). Average BOLD signal for all subjects is plotted against the scan number. The coloured regions depict the mean BOLD signal +/- 1 standard deviation.}
  	\label{fig:boldroi}
\end{figure}

@@

where $C_i$ denotes the probability of the class---sentence or picture. Two approaches were used to estimate the parameters of the model. In the first method, averages of the selected scans belonging to a stimulus were used to estimate the parameters, while in the second method, the parameters were estimated for each scan separately. In both methods the decision of the classifier is given by Eq.~\ref{eq:maxLikelihood}, which is the maximum likelihood of the two classes over all scans.

@@

\textbf{\emph{Linear Support Vector Machine (LSVM).}} The LSVM is a discriminant based approach to classification. For an elaborate description see \cite{bk:2004aplay}. For each selected scan belonging to a stimulus, the LSVM classifies it as either a picture or a sentence. The final classification of the stimulus is:

@@

\textbf{\emph{Multivariate Gaussian (MVG).}} In MVG, independence between feature vectors is not assumed. Instead, the multivariate distribution is used to estimate the probability of the classes, with the covariance matrix and the means as parameters. The log-probability of a class becomes:
\begin{dmath}
\label{eq:multivariate}
g_{i}(\mathbf{x}) = -\frac{d}{2}\log{2\pi}-\frac{1}{2}\log{|\Sigma_{i}|}-\frac{1}{2}(\mathbf{x}-\mu_{i})^{T}\Sigma_{i}^{-1}(\textbf{x}-\mu_{i})+\log{P(C_{i})}
\end{dmath}

>>>>

%% The ``\maketitle'' command must be the first command after the
%% ``\begin{document}'' command. It prepares and prints the title block.
%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}
\label{sec:introduction}

\maketitle

%% Now the actual contents of first section
Human functional brain mapping is a field of study that has grown strongly in the past decades. This is not a surprise, because it is one of the few ways to look at brain activity of a living animal. There are several techniques available to scan for brain activity in a living human. Of these techniques, functional magnetic resonance imaging (fMRI) has become more and more popular because it has a very good spatial resolution (contrary to MEG/EEG), and it is non-invasive (contrary to PET). Brain activity is inferred from differences in the ratio between oxygenated and deoxygenated blood; when brain activity occurs, more oxygenated blood will flow to a brain region, and this change is detected by a machine. This is called the blood oxygen level dependent (BOLD) response \cite{sm:2012fMRI}.

@@

Traditionally, data from each voxel, or from a group of voxels, are analyzed with analysis of variance. Voxels are usually convoluted with a hemodynamic response function in an attempt to infer the spiking neural activity from the slowly building and dropping BOLD response \cite{he:2002na} (cf.~Figure~\ref{fig:bold_response}). In this setup, experimental conditions are the independent variables and a general linear model is fitted to each voxel, or group of voxels. By using this technique, voxels or brain areas can be identified that differ in their average BOLD response across conditions.

@@

\subsection{Dimensionality Reduction}
\label{sec:dimensionalityReduction}
%Useful background info: http://www.brainvoyager.com/bvqx/doc/UsersGuide/WebHelp/Content/MVPATools/MVPA_Basic_Concepts.htm

Typical input for the machine learning algorithms discussed in this paper consists of a number of feature vectors, one for each fMRI scan. Such a feature vector $\vec{x}$ contains measures for the $N$ features we use to represent that scan and is derived from our input data. In the most basic case, we could use the entire set of voxels belonging to a scan, resulting in a feature vector containing potentially (hundreds of) thousands of features.

In the case that we have a very large dimension of features combined with only a limited amount of scans, our classifier may have difficulty learning the right classification function. This problem is formally known as the \emph{Hughes effect}\cite{Hughes1054102} which states that the predictive power of a machine learning algorithm decreases as the dimensionality increases with a fixed number of training examples. To reduce the dimensionality of the input, several different methods were used.

@@

\textbf{\emph{ROI mapping.}} ROI mapping is a form of feature selection, wherein only a subset of the N-dimensional input space is used. As described in the Section~\ref{sec:introduction}, ROIs are anatomically defined brain areas classified as relevant by an expert in the field. We used 7 of the 25 defined ROIs, labeled  'CALC' 'LIPL' 'LT' 'LTRIA' 'LOPER' 'LIPS' and  'LDLPFC' (cf.~Figure~\ref{fig:fd}). These ROIs have previously been found to be most relevant in discriminating between the cognitive states of looking at sentences or pictures \cite{wa:2003betw}. To further reduce the number of features, average voxel values within these regions are taken as new input features. As with preprocessing, other higher-order features such as the standard deviation, kurtosis, and skewness within ROIs did not increase predictability and were not included in the subsequent analysis. The disadvantage of using ROIs is that they require \emph{a priori} knowledge. When this knowledge is inaccurate, classification accuracy may not be optimal.

@@

\textbf{\emph{Temporal dimensionality reduction.}} The BOLD response peaks at approximately 5 seconds after a stimulus is presented (cf.~Figure~\ref{fig:bold_response}). This means that the first scans after a stimulus may not represent activity related to this stimulus and should not be used for classification. To examine which scans provide maximal class information, we plotted the average BOLD activity per scan for all trials of all subjects, given a trial type and ROI. As illustrated in Figure~\ref{fig:boldroi}, scans 10 to 20 provide maximal information for the first stimulus, while scans 23 to 33 provide maximal information for the second stimulus. The other scans were excluded.

\begin{figure}
	\centering
	\includegraphics[width=90mm]{figures/scan_hemodynamic_BOLD_roi}
  	\caption{Haemodynamic response for ROI "CALC". The red line depicts a sentence-picture trial (S-P) and the blue line depicts a picture-sentence trial (P-S). Average BOLD signal for all subjects is plotted against the scan number. The coloured regions depict the mean BOLD signal +/- 1 standard deviation.}
  	\label{fig:boldroi}
\end{figure}

@@

where $C_i$ denotes the probability of the class (picture or sentence). Two approaches were used to estimate the parameters of the model. In the first method, averages of the selected scans belonging to a stimulus were used to estimate the parameters, while in the second method, the parameters were estimated for each scan separately. In both methods the decision of the classifier is given by Eq.~\ref{eq:maxLikelihood}, which is the maximum likelihood of the two classes over all scans.

@@

\textbf{\emph{Linear Support Vector Machine (LSVM).}} The LSVM is a discriminant based approach to classification. For an elaborate description see Chapter 10 of \cite{bk:2004aplay}. For each selected scan belonging to a stimulus, the LSVM classifies it as either a picture or a sentence. The final classification of the stimulus is:

@@

\textbf{\emph{Multivariate Gaussian (MVG).}} In MVG, independence between feature vectors is not assumed. Instead, the multivariate distribution is used to estimate the probability of the classes, with the covariance matrix and the means as parameters. The log-probability of a class becomes:

\begin{dmath}
\label{eq:multivariate}
g_{i}(\mathbf{x}) = -\frac{d}{2}\log{2\pi}-\frac{1}{2}\log{|\Sigma_{i}|}-\frac{1}{2}(\mathbf{x}-\mu_{i})^{T}\Sigma_{i}^{-1}(\textbf{x}-\mu_{i})+\log{P(C_{i})}
\end{dmath}
